---
title: "Reproducible Research: Peer Assessment 1"
author: "Karen Linden"
date: "Sunday, June 14, 2015"
output: 
  html_document:
    keep_md: true
---
    
---

opts_chunk$set(echo=TRUE, results='asis')

## Loading and preprocessing the data

Read in the data.

```{r}
setwd('~/R/ReproducibleResearchProject1')

if(!file.exists('activity.csv')){
        unzip('repdata-data-activity.zip')
}

data.file <- read.csv("activity.csv") # Read in data file
```

Create a date.time column that combines the date and interval columns. Ended up not needs this field

```{r}
time.fix <- formatC(data.file$interval / 100, 2, format='f')
data.file$date.time <- as.POSIXct(paste(data.file$date, time.fix),
                                 format='%Y-%m-%d %H.%M',
                                 tz='GMT')
```

For analyzing the convenient date file To do this it will help us with the analysis.

```{r}
date.fix <- strptime(data.file$date, "%Y-%m-%d")
data.file$date.update <- date.fix
```

unique.dates and unique.intervals, they are variables that store a list of all possible dates and intervals. These will primarily be used to help plot our necessary data and for data processing.

```{r}
unique.dates <- unique(date.fix)
unique.intervals <- unique(data.file$interval)
```

## What is mean total number of steps taken per day?

First, calculate the mean number of steps for each day:

```{r}
steps.by.day <- split(data.file$steps, date.fix$yday)
steps.per.day <- sapply(steps.by.day, sum, na.rm=TRUE)

```
Mean and median for the total steps per day:

```{r}
mean.stepsperday <- sapply(steps.by.day, mean, na.rm=TRUE)
mean.dataframe <- data.frame(date=unique.dates, mean.stepsperday=mean.stepsperday, row.names=NULL)
mean.dataframe

median.stepsperday <- sapply(steps.by.day, median, na.rm=TRUE)
median.dataframe <- data.frame(date=unique.dates, median.stepsperday=median.stepsperday, row.names=NULL)
median.dataframe

```
Look at the distribution of total number of steps per day with a histogram:

```{r}

plot(unique.dates, 
     steps.per.day, 
     main="Histogram of Steps Taken Each Day", 
     xlab="Date (10-01-2012 Thru 11-30-2012)", 
     ylab="Steps", 
     type="h", 
     lwd=4, 
     col="red")
```

## What is the average daily activity pattern?

Calculate the mean steps for each five minute interval, and then put it in a data frame.

```{r}
interval.split <- split(data.file$steps, data.file$interval)

average.interval <- sapply(interval.split, mean, na.rm=TRUE)
```

Let's take a look at a time series plot for the mean steps.

```{r}
plot(unique.intervals, average.interval, 
     type="l",
     main="Average Number of Steps per Interval Across All Days", 
     xlab="Interval", 
     ylab="Average # of Steps Across All Days", 
     lwd=2, 
     col="green")

max.interval.day <- max(average.interval, na.rm=TRUE)
max.index <- as.numeric(which(average.interval == max.interval.day))


max.Interval <- unique.intervals[max.index]
abline(v=max.Interval , col="orange", lwd=3)

```
The maximum number of steps averaged across all days is:

```{r}
max.Interval
```

## Imputing missing values

Identify the number of intervals with missing step counts ("NA's"):

```{r}
missing.steps <- is.na(data.file$steps)
# How many missing
table(missing.steps)
```

To fill in the missing values, I'll use mean steps for a five-minute interval for the entire dataset.

```{r}
library(Hmisc)
averages <- aggregate(x=list(steps=data.file$steps), by=list(interval=data.file$interval),
                                             FUN=mean, na.rm=TRUE)

fill.value <- function(steps, interval) {
        filled <- NA
        if (!is.na(steps))
                filled <- c(steps)
        else
                filled <- (averages[averages$interval==interval, "steps"])
        return(filled)
}
fill.data <- data.file
fill.data$steps <- mapply(fill.value, fill.data$steps, fill.data$interval)

```

Compare the mean and median steps for each day between the original data set and the imputed data set.

```{r}
newsteps.by.day <- split(fill.data$steps, date.fix$yday)
newsteps.per.day <- sapply(newsteps.by.day, sum)
```

And a histogram of the imputed dataset.
```{r}
par(mfcol=c(2,1))


plot(unique.dates, 
     steps.per.day, 
     main="Histogram of Steps Taken Each Day", 
     xlab="Date (10-01-2012 Thru 11-30-2012)", 
     ylab="Steps", 
     type="h", 
     lwd=4, 
     col="red")

# Plot the modified histogram after

plot(unique.dates, 
     newsteps.per.day, 
     main="Histogram of Steps Taken Each Day", 
     xlab="Date (10-01-2012 Thru 11-30-2012)", 
     ylab="Steps", 
     type="h", 
     lwd=4, 
     col="red")
```

Mean and median for the total steps per day:

```{r}
newmean.stepsperday <- sapply(newsteps.by.day, mean, na.rm=TRUE)
newmean.dataframe <- data.frame(date=unique.dates, newmean.stepsperday=newmean.stepsperday, row.names=NULL)
newmean.dataframe

newmedian.stepsperday <- sapply(newsteps.by.day, median, na.rm=TRUE)
newmedian.dataframe <- data.frame(date=unique.dates, newmedian.stepsperday=newmedian.stepsperday, row.names=NULL)
newmedian.dataframe
```

Imputing the missing data has increased the average number of steps.


## Are there differences in activity patterns between weekdays and weekends?

Add a column for whether a day is a weekday or weekend.

```{r}
dayofweek <- function(date) {
        day <- weekdays(date)
        if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
                return("weekday")
        else if (day %in% c("Saturday", "Sunday"))
                return("weekend")
        else
                stop("invalid date")
}

fill.date <- fill.data 
fill.date$date <- as.Date(fill.data$date)
fill.date$dayofweek <- sapply(fill.date$date, FUN=dayofweek)
```


```{r}
weekday <- fill.date[fill.date$dayofweek == "weekday", ]
weekend <- fill.date[fill.date$dayofweek == "weekend", ]

weekday.split <- split(weekday$steps, weekday$interval)
weekend.split <- split(weekend$steps, weekend$interval)

mean.weekday.split <- sapply(weekday.split, mean)
mean.weekend.split <- sapply(weekend.split, mean)
```
Compare the patterns between weekdays and weekends.

```{r}
par(mfcol=c(2,1))

plot(unique.intervals, 
     mean.weekday.split, 
     type="l",
     main="Average number of steps per interval across all weekdays", 
     xlab="Interval", 
     ylab="# of Steps on weekdays", 
     lwd=2, 
     col="red")

plot(unique.intervals, 
     mean.weekend.split, 
     type="l",
     main="Average number of steps per interval across all weekends", 
     xlab="Interval", 
     ylab="# of steps om weekends", 
     lwd=2, 
     col="red")
```